Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-19 paper/paper.tex
fix typos

In this work, we present the design of \theoryname (\theoryabbv), a
domain-specific language for formal verification of parsers. A key
property is that \theoryabbv is an \emph{extensible} framework for
verification of parsers in that it supports the definition of grammar
formalisms of unrestricted complexity. That is, \theoryabbv is not a
system for verifying \emph{one} type of grammar formalism, but instead
is a domain-specific language in which many grammar formalisms and
their verified parsers can be implemented. For example, \theoryabbv is
not a verified parser generator that compiles regular expressions to
deterministic finite automata, but is instead is a domain specific
language \emph{for writing} such a verified parser generator.

@@

\paragraph{Finite Grammars}
First consider finite grammars --- those built from base types via disjunctions and
concatenations. The base types comprise characters drawn from the alphabet, the
empty string, and the empty grammar.
For each character $a$ in the alphabet we have a type $\literal a$ which
has a single parse tree for the string
\stringquote{a} and no parse trees at any other strings. The grammar $\I$ has a single
parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
The final base type, the empty grammar $0$, has no parses for any string. We
use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
$\otimes$ of
grammars. Over an input string $\w$, a parse of the disjunction $\A \oplus \B$ is either
a parse of $\A$ over the string $\w$ or a
parse of $\B$ over the string $\w$. Similarly, $\w$ matches $\A \otimes \B$ if
$\w$ can be split into two strings $\w_{\A}$ and $\w_{\B}$ that match $\A$
and $\B$, respectively.

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
f (a , b) = inl (a (*@$\otimes$@*) b)
\end{lstlisting}\begin{mathpar}
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash a \otimes b : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a \otimes b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{\stringquote{ac} matches
  $(\literal a \oplus \literal b) \otimes \literal c$}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\[
  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
\]

@@

Through repeated application of the Kleene star constructors,
\cref{fig:kleenestarderivation} gives a derivation that shows
\stringquote{ab} matches the regular expression $({\literal a}^{*}
\otimes \literal b) \oplus \literal c$. The leaves of the proof tree
that mention the arrow $\uparrow$ describe a cast from a non-linear
type to a linear type.  For instance, the premise of the leaf
involving $\nil$ views $\nil : \ltonl {({\literal a}^{*})}$ as the
name of a constructor, and a constructor should be nonlinearly valued
because we may call it several times (or not at all). However, the
conclusion of this leaf views $\nil : {\literal a}^{*}$ as a linear
value, which in our syntax is an implicit coercion from a nonlinear
value to a linear value. After we call the constructor it ``returns''
a value that may only be used a single time.

\begin{figure}
\begin{mathpar}
\footnotesize
\inferrule
{
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {
          \inferrule
          {~}
          {\cdot \vdash \cons : \ltonl {({\literal a} \lto {\literal a}^* \lto {\literal a}^*)}}
        }
        {\cdot \vdash \cons : {\literal a} \lto {\literal a}^* \lto {\literal a}^*}
        \\
        \inferrule
        {~}
        {a : {\literal a} \vdash a : {\literal a}}
      }
      {a : a \vdash \cons~a : {\literal a}^* \lto {\literal a}^*}
      \\
      \inferrule
      {
        \inferrule
        {~}
        {\cdot \vdash \nil : \ltonl {({\literal a}^*)}}
      }
      {\cdot \vdash \nil : {\literal a}^*}
    }
    {a : {\literal a} \vdash \cons~a~\nil : {\literal a}^{*}}
    \\
    \inferrule
    {~}
    {b : {\literal b} \vdash b : {\literal b}}
  }
  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil) \otimes b : {\literal a}^{*} \otimes {\literal b}}
}
{a : \literal a , b : \literal b \vdash \texttt{g} := \inl ((\cons~a~\nil)
  \otimes b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
\end{mathpar}
\begin{lstlisting}
g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
g (a , b) = inl (cons a' nil (*@$\otimes$@*) b')
\end{lstlisting}
\caption{\stringquote{ab} matches
  $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
\label{fig:kleenestarderivation}
\end{figure}

@@

\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons (a_1 , \cons(a_2 , a_{*})) : \A^*}
      }
      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2} {a'} {\cons (a_1 , \cons(a_2 , a_{*})) : \A^*}}
    }
    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

First, we cover the ``multiplicative'' connectives of non-commutative
linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
are standard for a non-commutative linear logic: when we construct a
linear unit we cannot use any variables and when we construct a tensor
product, the two sides must use disjoint variables, and the variables
the left side of the product uses must be to the left in the context
of the variables used by the right side of the tensor product. The
elimination rules for unit and tensor are given by pattern
matching. The pattern matching rules split the linear context into
three pieces $\Delta_1,\Delta_2,\Delta_3$: the middle $\Delta_2$ is
used by the scrutinee of the pattern match, and in the continuation
this context is replaced by the variables brought into scope by the
pattern match. This ensures that pattern matches maintain the proper
ordering of resource usage.

@@

\begin{figure}
  \begin{footnotesize}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
  \end{mathpar}
\begin{mathpar}
    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta' \vdash e' : A \\ \Gamma ; \Delta \vdash e : A \lto B}{\Gamma ; \Delta, \Delta' \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}    
  \end{footnotesize}
  \caption{Linear types and terms (selection)}
  \label{fig:linear-types-and-terms}
\end{figure}

@@

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: up to weak generative capacity, meaning just which strings are
accepted by the grammar; and up to \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.
\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exists
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parse of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to match the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of our axiom that $\oplus$ constructors are
disjoint.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

A regular expressions in \theoryabbv is a linear type constructed
using only the connectives $\literal c$, $0$, $\oplus$, $I$,
$\otimes$, and Kleene star.  In \cref{sec:type-theory-examples}, we
saw one particular NFA and its corresponding type of traces. More
generally we define the linear type of traces as in
\Cref{fig:nfatrace}

@@

Next, we give two examples for parsing context-free
grammars. Context-free grammars (CFG) can be encoded in our type
theory in a similar way to regular expressions, as CFGs are equivalent
to the formalism of \emph{$\mu$}-regular expressions, where the Kleene
star is replaced by an arbitrary fixed point
operation\cite{leis_towards_1992}.

>>>>

In this work, we present the design of \theoryname (\theoryabbv), a
domain-specific language for formal verification of parsers. A key
property is that \theoryabbv is an \emph{extensible} framework for
verification of parsers in that it supports the definition of grammar
formalisms of unrestricted complexity. That is, \theoryabbv is not a
system for verifying \emph{one} type of grammar formalism, but instead
is a domain-specific language in which many grammar formalisms and
their verified parsers can be implemented. For example, \theoryabbv is
not a verified parser generator that compiles regular expressions to
deterministic finite automata, but is instead a domain specific
language \emph{for writing} such a verified parser generator.

@@

\paragraph{Finite Grammars}
First consider finite grammars --- those built from base types via disjunctions and
concatenations. The base types comprise characters drawn from the alphabet, the
empty string, and the empty grammar.
For each character $a$ in the alphabet we have a type $\literal a$ which
has a single parse tree for the string
\stringquote{a} and no parse trees at any other strings. The grammar $\I$ has a single
parse tree for the empty string $\epsilon = \stringquote{}$ and no parses for any other strings.
The final base type, the empty grammar $0$, has no parses for any string. We
use type-theoretic syntax to represent disjunction $\oplus$ and concatenation
$\otimes$ of
grammars. Over an input string $\w$, a parse of the disjunction $\A \oplus \B$ is either
a parse of $\A$ over the string $\w$ or a
parse of $\B$ over the string $\w$. Similarly, a parse of $\A \otimes \B$ for
$\w$ is a splitting of $w$ into two strings $\w_{\A}$ and $\w_{\B}$ with
parses for $\A$
and $\B$, respectively.

@@

\begin{figure}
\begin{lstlisting}
f : (*@$\uparrow$@*)('a' (*@$\otimes$@*) 'b' (*@$\lto$@*) ('a' (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
f (a , b) = inl (a (*@$\otimes$@*) b)
\end{lstlisting}\begin{mathpar}
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {~}
      {a : \literal a \vdash a : \literal a}
      \\
      \inferrule
      {~}
      {b : \literal b \vdash b : \literal b}
    }
    {a : \literal a , b : \literal b \vdash (a , b) : \literal a \otimes
      \literal b}
  }
  {a : \literal a , b : \literal b \vdash \texttt{f} := \inl(a , b) :
    (\literal a \otimes \literal b) \oplus \literal c}
\end{mathpar}
\caption{\stringquote{ab} is parsed by $(\literal a \oplus \literal b) \otimes \literal c$
}
\label{fig:fingram}
\end{figure}
For this interpretation of parse trees as terms to make sense, our
calculus cannot allow for \emph{any} of the usual structural rules of
type theory: weakening, contraction and exchange. Weakening allows
for variables to go unused, while contraction allows for the same
variable to be used twice, but in a parse tree, every character must
be accounted for exactly once. That is, we want to prevent the following
erroneous derivations,
\[
  a : \literal a , b : \literal b \not \vdash a : \literal a \qquad a : \literal a , b : \literal b \not \vdash (a, a) : \literal a \otimes \literal a
\]

@@

Through repeated application of the Kleene star constructors,
\cref{fig:kleenestarderivation} gives a derivation that shows
\stringquote{ab} is parsed by the regular expression $({\literal a}^{*}
\otimes \literal b) \oplus \literal c$. The leaves of the proof tree
that mention the arrow $\uparrow$ describe a cast from a non-linear
type to a linear type.  For instance, the premise of the leaf
involving $\nil$ views $\nil : \ltonl {({\literal a}^{*})}$ as the
name of a constructor, and a constructor should be nonlinearly valued
because we may call it several times (or not at all). However, the
conclusion of this leaf views $\nil : {\literal a}^{*}$ as a linear
value, which in our syntax is an implicit coercion from a nonlinear
value to a linear value. After we call the constructor it ``returns''
a value that may only be used a single time.

\begin{figure}
\begin{mathpar}
\footnotesize
\inferrule
{
  \inferrule
  {
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {
          \inferrule
          {~}
          {\cdot \vdash \cons : \ltonl {({\literal a} \lto {\literal a}^* \lto {\literal a}^*)}}
        }
        {\cdot \vdash \cons : {\literal a} \lto {\literal a}^* \lto {\literal a}^*}
        \\
        \inferrule
        {~}
        {a : {\literal a} \vdash a : {\literal a}}
      }
      {a : a \vdash \cons~a : {\literal a}^* \lto {\literal a}^*}
      \\
      \inferrule
      {
        \inferrule
        {~}
        {\cdot \vdash \nil : \ltonl {({\literal a}^*)}}
      }
      {\cdot \vdash \nil : {\literal a}^*}
    }
    {a : {\literal a} \vdash \cons~a~\nil : {\literal a}^{*}}
    \\
    \inferrule
    {~}
    {b : {\literal b} \vdash b : {\literal b}}
  }
  {a : {\literal a} , b : {\literal b} \vdash (\cons~a~\nil , b) : {\literal a}^{*} \otimes {\literal b}}
}
{a : \literal a , b : \literal b \vdash \texttt{g} := \inl (\cons~a~\nil , b) : ({\literal a}^{*} \otimes \literal b) \oplus \literal c}
\end{mathpar}
\begin{lstlisting}
g : (*@$\uparrow$@*)(('a' (*@$\otimes$@*) 'b') (*@$\lto$@*) ('a'(*@$^*$@*) (*@$\otimes$@*) 'b') (*@$\oplus$@*) 'c')
g (a , b) = inl (cons a nil (*@$\otimes$@*) b)
\end{lstlisting}
\caption{\stringquote{ab} is parsed by
  $({\literal a}^{*} \otimes \literal b) \oplus \literal c$}
\label{fig:kleenestarderivation}
\end{figure}

@@

\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
      }
      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2}
        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
    }
    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 (*@$\otimes$@*) a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

Next, we cover the ``multiplicative'' connectives of non-commutative
linear logic. The linear unit ($\I$) and tensor product ($\otimes$)
are standard for a non-commutative linear logic: when we construct a
linear unit we cannot use any variables and when we construct a tensor
product, the two sides must use disjoint variables, and the variables
the left side of the product uses must be to the left in the context
of the variables used by the right side of the tensor product. The
elimination rules for unit and tensor are given by pattern
matching. The pattern matching rules split the linear context into
three pieces $\Delta_1,\Delta_2,\Delta_3$: the middle $\Delta_2$ is
used by the scrutinee of the pattern match, and in the continuation
this context is replaced by the variables brought into scope by the
pattern match. This ensures that pattern matches maintain the proper
ordering of resource usage.

@@

\begin{figure}
  \begin{footnotesize}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash \I \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \otimes B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \lto B \isLinTy}\and
    \inferrule{\Gamma \vdash A \isLinTy \and \Gamma \vdash B \isLinTy}{\Gamma \vdash A \tol B \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigoplus\limits_{x:X} A \isLinTy}\and
    \inferrule{\Gamma,x:X \vdash A \isLinTy}{\Gamma \vdash \bigamp\limits_{x:X} A \isLinTy}\and
    \inferrule
    {\Gamma \vdash f : \ltonl {(A \lto B)} \and \Gamma \vdash g : \ltonl { (A \lto B) }}
    {\Gamma \vdash \equalizer {a}{f}{g} \isLinTy}
  \end{mathpar}
\begin{mathpar}
    \inferrule{~}{\Gamma ; a : A \vdash a : A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma ; \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma ; \cdot \vdash () : I}
    \and
    \inferrule{\Gamma ; \Delta_2 \vdash e : I \\ \Gamma ; \Delta_1,\Delta_3 \vdash e' : C}{\Gamma ; \Delta_1,\Delta_2,\Delta_3 \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma ; \Delta \vdash e : A \\ \Gamma ; \Delta' \vdash e' : B}{\Gamma ; \Delta, \Delta' \vdash (e , e') : A \otimes B}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : A \otimes B \\ \Gamma ; \Delta_1, a : A, b : B, \Delta_2 \vdash e' : C}{\Gamma ;  \Delta_1, \Delta_2, \Delta_3 \vdash \letin {(a , b)} e {e'} : C}
    \\
    %
    \inferrule{\Gamma ; \Delta , a : A \vdash e : B}{\Gamma ; \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : A \lto B \\ \Gamma ; \Delta' \vdash e' : A} {\Gamma ; \Delta, \Delta' \vdash \applto {e} {e'} : B}
    \\
    %
    \inferrule{\Gamma, x : X ; \Delta  \vdash e : A}
              {\Gamma ; \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma ; \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash M : X}{\Gamma ; \Delta \vdash e\,.\pi\,M : \subst A {M} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash M : X \quad \Gamma ; \Delta \vdash e : \subst A M x}{\Gamma ; \Delta \vdash \sigma\,M\,e : \bigoplus\limits_{x:X} A}
    %
    \and
    %
    \inferrule{\Gamma ; \Delta_2 \vdash e : \bigoplus\limits_{x:X} A \quad \Gamma, x : X ; \Delta_1, a : A, \Delta_3 \vdash e' : C}{\Gamma; \Delta_1, \Delta_2, \Delta_3 \vdash \letin {\sigma\,x\,a} e {e'}: C}
    %
    \and
    %
    \inferrule
    { \Gamma ; \Delta \vdash e : A \\
      \Gamma ; \Delta \vdash \applto {f}{e} \equiv \applto {g}{e}}
    {\Gamma ; \Delta \vdash \equalizerin{e} : \equalizer {a}{f}{g}}
    %
    \and
    %
    \inferrule
    {\Gamma ; \Delta \vdash e : \equalizer{a}{f}{g}}
    {\Gamma ; \Delta \vdash \equalizerpi {e} : A}
  \end{mathpar}    
  \end{footnotesize}
  \caption{Linear types and terms (selection)}
  \label{fig:linear-types-and-terms}
\end{figure}

@@

%% \paragraph{Weak and Strong Equivalence}
In the theory of formal grammars, there are two different notions of
equivalence: up to weak generative capacity, meaning just which strings are
accepted by the grammar; and up to \emph{strong} generative capacity, when
the parse trees of the two grammars are isomorphic
\cite{chom1963}. Using linear types as grammars, we can define both of
these notions of equivalence in \theoryabbv.
\begin{definition}
  \label{def:weakequiv}
  Grammars $\A$ and $\B$ are \emph{weakly equivalent} if there exist
  parse transformers $\f : \ltonl{(\A \lto \B)}$ and $\g : \ltonl {(\B
    \lto \A)}$. $\A$ is a \emph{retract} of $\B$ if $\A$ and $\B$ are
  weakly equivalent and $\lambda a. g(f(a)) \equiv \lambda a.a$. They
  are \emph{strongly equivalent} if further the other composition is
  the identity, i.e., $\lambda b. f(g(b)) \equiv \lambda b.b$.
\end{definition}

@@

%% \paragraph{Ambiguity}
A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory\cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\agdalogo]
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
  If a disjunction $\bigoplus\limits_{x:X} A(x)$ is unambiguous then
  each $A(x)$ is unambiguous.
\end{lemma}
From the first principle, we can prove that $\String$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\String$ is also strongly equivalent to $\top$. Next, since $\String$ is
defined as a Kleene star, we can easily show that $\String \cong \I
\oplus \Char \oplus (\Char \otimes \Char \otimes \String)$. Then by
the second principle we have that $\I$ and $\Char$ and each literal
$\literal c$ are unambiguous as well.

@@

We now turn to our main task, which is using our linear type system to
implement verified parsers. Given a grammar defined as a linear type
$A$, a first attempt at defining a parser would be to implement a
function $\uparrow{(\String \lto A)}$. But since our linear functions
must be total, this means that we can construct an $A$ parse for
\emph{every} input string, which is impossible for most grammars of
interest. Instead we might try to write a partial function as a
$\ltonl{(\String \lto (A \oplus \top))}$ using the ``option''
monad. This allows for the possibility that the input string doesn't
parse, but is far too weak as a specification: we can trivially
implement a parser for any type by always returning $\inr$. The
correct notion of a parser should be one that allows for failure, but
only in the case that a parse cannot be constructed.
\begin{definition}
  A parser for a linear type $A$ is a function $\uparrow{(\String \lto
    A \oplus A_{\neg})}$, where $A_{\neg}$ is a linear type that is
  \emph{disjoint} from $A$ in that we can implement a function
  $\uparrow{(A \& A_{\neg} \lto 0)}$.
\end{definition}
Here we replace $\top$ in our partial parser type with a type
$A_{\neg}$ that we can think of as a negation of $A$. The function
$\uparrow{(A \& A_{\neg} \lto 0)}$ ensures that it is impossible for
$A$ and $A_{\neg}$ to parse the same input string. This means that in
defining a parser, we will need to define a kind of negative grammar
for strings that do not parse. Fortunately, we will see that
deterministic automata naturally support such a notion with no
additional effort: the negative grammar is simply the grammar for
traces that end in a rejecting state. This follows from the following
principle, a consequence of our axiom that $\oplus$ constructors are
disjoint.
\begin{lemma}[\agdalogo]
  If $A \oplus A_{\neg}$ is unambiguous, then $A$ and $A_{\neg}$ are
  disjoint.
\end{lemma}

@@

A regular expression in \theoryabbv is a linear type constructed
using only the connectives $\literal c$, $0$, $\oplus$, $I$,
$\otimes$, and Kleene star.  In \cref{sec:type-theory-examples}, we
saw one particular NFA and its corresponding type of traces. More
generally we define the linear type of traces as in
\Cref{fig:nfatrace}

@@

Next, we give two examples for parsing context-free
grammars. Context-free grammars (CFG) can be encoded in our type
theory in a similar way to regular expressions, as CFGs are equivalent
to the formalism of \emph{$\mu$}-regular expressions, where the Kleene
star is replaced by an arbitrary fixed point
operation \cite{leis_towards_1992}.
